Review on precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-04 manuscript.tex
importing from manubot : cleanup

\hypertarget{introduction-importance-of-precise-spike-timings-in-the-brain}{%
\section{Introduction: Importance of precise spike timings in the brain}\label{introduction-importance-of-precise-spike-timings-in-the-brain}}

\hypertarget{ultra-fast-neural-codes-for-ultra-fast-vision}{%
\subsection{Ultra fast neural codes for ultra fast vision}\label{ultra-fast-neural-codes-for-ultra-fast-vision}}

@@

\begin{figure}
\hypertarget{fig:thorpe}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{https://cdn-learn.adafruit.com/assets/assets/000/086/572/original/circuitpython_Thorpe-FabreThorpe-Neuroscience-Seeking-categories-in-the-brain-monkey-visual-latency-estimate.jpg}
\caption{From {[}\protect\hyperlink{ref-12dWDpY61}{6}{]}}\label{fig:thorpe}
}
\end{figure}

\hypertarget{how-timing-encodes-analogous-profile}{%
\subsection{How timing encodes analogous profile}\label{how-timing-encodes-analogous-profile}}

An important characteristic of neuronal information is that it consists mainly in the transfer of action potentials, or spikes, which consist of brief impulses that propagate along the axons of neurons. These have the particularity of being essentially binary in their amplitude (that is to say, they are prototypical, all or nothing). An important consequence of the speed of processing is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only 120 ms consists of about ten processing stages following the ``forward'' pathways of the visual system, then this imposes that the processing in a single area is performed with a reduced number of spikes. Neurophysiologists tipcally use firing sequences to characterize the activity of neurons using different statistics on their individual timing {[}\protect\hyperlink{ref-zZUVkTfZ}{7}{]} but also the dependance across neurons {[}\protect\hyperlink{ref-13T2Gne9z}{8}{]}.

@@

\begin{figure}
\hypertarget{fig:roc}{%
\centering
\includegraphics{images/roc.jpg}
\caption{Latency coding. An input analog profile is encoded in latencies: the higher the contrast, the shorter the latency. In this example, one generates at most one spike per neuron. From {[}\protect\hyperlink{ref-YX5wqWH4}{16}{]}}\label{fig:roc}
}
\end{figure}

@@

\hypertarget{from-synfire-chains}{%
\subsection{From synfire chains\ldots{}}\label{from-synfire-chains}}

The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate is highly irregular {[}\protect\hyperlink{ref-sYHyiGsP}{26}{]}, which makes it inconsistent with temporal integration of random EPSPs. It has also been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible, where the first spike is aligned but the subsequent spike times tend to diffuse for independent repetitions of the stimulation {[}\protect\hyperlink{ref-XsHCA4IZ}{15}{]} (see Figure {[}\ref{fig:mainen}{]}). When driven by a highly dynamic \emph{frozen} noise, the output spikes are highly reproducible. This is consistent with the differential role of different stimulus frequencies on the reliability of spike timing reported in {[}\protect\hyperlink{ref-1AbvCTG03}{27}{]} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''

\begin{figure}
\hypertarget{fig:mainen}{%
\centering
\includegraphics{https://camo.githubusercontent.com/77f4a48a0fe8f98ffab3c1f21dca1e74286b14d47c17120ee9627f3351fcd5e7/687474703a2f2f692e737461636b2e696d6775722e636f6d2f69786e727a2e706e67}
\caption{The timing of the spikes produced following the repetition of a step stimuli is less reproducible than that to a noisy stimulus. While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} nois at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this \href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:diesman}{%
\centering
\includegraphics{https://brian2.readthedocs.io/en/stable/_images/frompapers.Diesmann_et_al_1999.1.png}
\caption{Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{using Brian}. The model consists of 10 groups of 100 neurons each. A pulse with a given jitter is generated in the first group which here generates a pulse after a certain processing delay in the second group with a lower jitter. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:pynn}{%
\centering
\includegraphics{https://www.frontiersin.org/files/Articles/39056/fnins-07-00011-r2/image_m/fnins-07-00011-g004.jpg}
\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background is only utilized in the original model, where it is implemented as random Gaussian current. For the presented hardware implementation it has been omitted due to network size constraints. As compensation for missing background stimuli, the resting potential was increased to ensure a comparable excitability of the neurons. The state space gives the duration of synfire chains as a function of the protocol's parameters and compares (C) CPU to (D) neuromorphic implementations.}\label{fig:pynn}
}
\end{figure}

@@

\hypertarget{to-traveling-waves}{%
\subsubsection{.. to traveling waves \ldots{}}\label{to-traveling-waves}}

TODO : {[}\protect\hyperlink{ref-zNDCo9Hy}{49}{]}

@@

\begin{figure}
\hypertarget{fig:davis}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{https://media.springernature.com/lw685/springer-static/image/art\%3A10.1038\%2Fs41467-021-26175-1/MediaObjects/41467_2021_26175_Fig1_HTML.png}
\caption{a Spike raster for repeated presentations (N = 40) of a high-contrast (10\% Michelson contrast) drifting Gabor recorded from area MT of a fixating marmoset (stimulus-onset, red line; mean response, blue line). A single-trial LFP trace is plotted in gray, and the average LFP response is plotted in black. The relative power between baseline (−200 ms to stimulus-onset) and evoked fluctuations (stimulus-onset + 50--250 ms) significantly favored the evoked response (right panel; N = 110 trials; median = 1.89 dB, p = 0.000019, two-tailed Wilcoxon's ranked-sum test). b Same as in (a), but for a low contrast stimulus (\textless2\% Michelson contrast). The relative power between baseline and evoked LFP fluctuations was not statistically different from parity (median = 1.23 dB, p = 0.087 two-tailed Wilcoxon's ranked-sum test). c An example of spontaneous LFP fluctuations structured as a traveling wave recorded from a spatially distributed multielectrode array in marmoset area MT. d Histogram of spontaneous spike probability as a function of the generalized phase of the LFP during fixation. e The average evoked response to low contrast stimuli was stronger when a more excitable phase (±π rad) of a spontaneous traveling wave aligned with the retinotopic location of the target (aligned, green line) as compared to when a less excitable phase (0 rad) was aligned (unaligned, purple line; N = 43 wave and non-wave trials; shaded region SEM; p = 0.0000015 two-tailed Wilcoxon's rank-sum test). Data for panels c--e modified from Davis et al.17 with permission.}\label{fig:davis}
}
\end{figure}

@@

\hypertarget{spike-synchronization-and-rate-modulation-differentially-involved-in-motor-cortical-function-doi10.1126science.278.5345.1950}{%
\subsubsection{\texorpdfstring{Spike Synchronization and Rate Modulation Differentially Involved in Motor Cortical Function {[}\protect\hyperlink{ref-wVZpQvSk}{42}{]}}{Spike Synchronization and Rate Modulation Differentially Involved in Motor Cortical Function {[}42{]}}}\label{spike-synchronization-and-rate-modulation-differentially-involved-in-motor-cortical-function-doi10.1126science.278.5345.1950}}

It is now commonly accepted that planning and execution of movements are based on distributed processing by neuronal populations in motor cortical areas {[}\protect\hyperlink{ref-yLVM9IXi}{57}{]}. It is less clear, though, how these populations organize dynamically to cope with the momentary computational demands. Simultaneously recorded activities of neurons in the primary motor cortex of monkeys during performance of a delayed-pointing task exhibited context-dependent, rapid changes in the patterns of coincident action potentials. Accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.

@@

\hypertarget{todo-and-to-synfire-braids-and-polychronous-groups}{%
\subsubsection{TODO \ldots{} and to synfire braids and polychronous groups}\label{todo-and-to-synfire-braids-and-polychronous-groups}}

\begin{itemize}
\tightlist
\item
  {[}\protect\hyperlink{ref-HjDb1gnq}{59}{]} : from synfire chains to Synfire braids
\end{itemize}

@@

\hypertarget{are-there-precise-heterosynaptic-spiking-motifs-in-the-brain}{%
\subsection{Are there precise heterosynaptic spiking motifs in the brain?}\label{are-there-precise-heterosynaptic-spiking-motifs-in-the-brain}}

Currently, there is a consensus for rate-based coding models in computational and biological neuroscience. Nevertheless, there is a substantial literature in neurobiology indicating that brain dynamics often organize into stereotyped sequences (like synfire chains {[}\protect\hyperlink{ref-Mf6h1KNY}{63}{]}, packets {[}\protect\hyperlink{ref-2GovoaYw}{64}{]} or hippocampal sequences {[}\protect\hyperlink{ref-SVfasyCp}{65},\protect\hyperlink{ref-PUNGVou}{66}{]} and on the role of such precise spike timing in downstream information transfer and coding {[}\protect\hyperlink{ref-19fAM6hcv}{67},\protect\hyperlink{ref-13dQ0R4UM}{68},\protect\hyperlink{ref-MfmcMsqA}{69}{]}. This is for instance relevant in sensory pathways in vision {[}\protect\hyperlink{ref-1EHZgcKyL}{70}{]}, audition {[}\protect\hyperlink{ref-FlKJw2VL}{71}{]}, olfaction {[}\protect\hyperlink{ref-AhSRXBJ2}{72}{]}{]} or touch {[}\protect\hyperlink{ref-gDinei4a}{73}{]}. In particular, one theoretical viewpoint considers synfire braids (Bienenstock, 1995), where a precise sequential motif of spikes will synchronize as it reaches the soma of a neuron for which synaptic delays are adequately tuned. In particular, computational modeling shows that at the scale of neurons, an efficient neural code can emerge where spike times are organized in prototypical, precise temporal motifs (Izhikevich, 2006) which he defined as polychronous groups.

@@

\begin{figure}
\hypertarget{fig:haimerl}{%
\centering
\includegraphics{https://www.pnas.org/cms/10.1073/pnas.1718518116/asset/ab9a48f6-b28f-40c1-978e-6f6dbec39b98/assets/graphic/pnas.1718518116fig01.jpeg}
\caption{Mixed distance and duration representation in CA1. (A) Calcium fluorescence (heatmap) of CA1 neurons participating to run sequences in consecutive imaging sessions. Cells have been selected and ordered with respect to their activity in the first imaging session (Top). The black line represents speed of the mouse.}\label{fig:haimerl}
}
\end{figure}

@@

\hypertarget{cortical-songs}{%
\subsection{Cortical songs}\label{cortical-songs}}

\begin{itemize}
\tightlist
\item
  Ikegaya and colleagues {[}\protect\hyperlink{ref-Mf6h1KNY}{63}{]} worked on spontaneous activity in vitro and in vivo. They demonstrated that in cortical activity, we can find a repetition of several motifs. In PSCs, but also in spike activity. These sequences repeat after minutes and have a precise spatio temporal structure with a ms precision. They can be specific of a particular layer or colomn, are synchronized with network activity oscillation and can involve several cells. They also demonstated that these sequences can form supersequences : the cortical songs. It consist of the assembly of several sequences which repeat in a specific order with a compressed timing.
\item
  ``We find precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons in vivo and in vitro. These patterns repeat after minutes, maintaining millisecond accuracy.''
\item
  Precisely repeating motifs of spontaneous synaptic activity in slices: duration around 1s +/- .5 s. Some events in motifs are of similar size but sometimes absent - better described by Bernouilli than SE (and covariance)
\item
  \emph{in vivo} spontaneous activity also reveals repeating sequences. About 3000 sequences, each involving 3-10 cells out of about 900, and last up to 3 seconds
\item
  topography: ``Sequences had specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster (Fig. 4, A and B, and fig.~S3B). (\ldots) Therefore, repeating temporal patterns of activation (\ldots) were associated with a structured spatial organization of the neurons that formed them.''
\item
  ``Cortical songs: modular assemblies of repeated sequences'': hierarchical detection.
\item
  in cotical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)
\end{itemize}

\begin{figure}
\hypertarget{fig:Ikegaya2004}{%
\centering
\includegraphics{images/Ikegaya2004zse0150424620001.jpeg}
\caption{Fig. 1. from {[}\protect\hyperlink{ref-Mf6h1KNY}{63}{]} Repeated motifs of spontaneous synaptic activity in vitro and in vivo. (A) Repeated motifs of intracellular activity from layer 5 pyramidal neurons in slices. Panels show segments (red) of the same voltage-clamp recording from the same cell repeating seconds or minutes after the initial occurrence (blue). Arrows indicate timings of repeated PSCs. (i) Upper trace: low--temporal resolution display of spontaneous activity of a neuron. Lower traces: higher resolution display of the repeated motif at indicated regions of the trace (a to c). (ii) Example of a longer motif. (B) Three repetitions of a motif. The top traces show the motifs superimposed on each other (blue, green, and red), the middle traces show these same traces individually, and the bottom traces show temporally magnified regions of the motifs (a to c). (C) Repeated sequences of intracellular current-clamp recordings in vivo. Two (i) and three (ii) repetitions of motifs are shown. Shuffle tests were performed on traces (i), a to c, yielding significantly fewer repeats (fig.~S2, P \textless{} 0.02). In (i), the blue trace is shifted --2.75 mV; in (ii), the blue trace is shifted --1.58 mV, and the green +0.79 mV.}\label{fig:Ikegaya2004}
}
\end{figure}

@@

\hypertarget{outline}{%
\subsection{outline}\label{outline}}

A natural candidate to use these precise temporal patterns in the brain is to use Spiking Neural Networks (SNNs) {[}\protect\hyperlink{ref-DUpFdm8C}{78}{]}. The approach which is at present most prominent in the SNNs community is to use existing algorithms from machine learning and to adapt them to the specificity of spiking architectures. One such example is to adapt the successes of deep learning algorithms and to transfer the back-propagation algorithm to SNNs, for instance with a surrogate gradient. This approach is quite successful, and SNNs approach in some case the performance of Deep Learning algorithms, for instance on the N-MNIST dataset for categorizing digits in a stream of events. However, most biological neural systems use spikes and are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption. There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs. Our approach will be to focus on the temporal representation of information directly. In particular, our objective is to fully exploit the capacity of spiking neurons to detect synchronous patterns.

@@

{]}\{.banner .lightred\}

@@

\begin{figure}
\hypertarget{fig:silicon_retina}{%
\centering
\includegraphics{https://laurentperrinet.github.io/grant/anr-anr/event_driven_computations.png}
\caption{A miniature, event-based ATIS sensor. Contrary to a classical frame-based camera for which a full dense image representation is given at discrete, regularly spaced timings, the event-based camera provides with events at the micro-second resolution. These are sparse as they represent luminance increments or decrements (ON and OFF events, respectively).}\label{fig:silicon_retina}
}
\end{figure}

@@

\hypertarget{theoretical-models-for-the-detection-of-precise-heterosynaptic-spiking-motifs}{%
\section{Theoretical models for the detection of precise heterosynaptic spiking motifs}\label{theoretical-models-for-the-detection-of-precise-heterosynaptic-spiking-motifs}}

\#sec:detection

@@

\hypertarget{izhikevitchs-polychronization-model}{%
\subsection{Izhikevitch's polychronization model}\label{izhikevitchs-polychronization-model}}

Considering the number of spikes required to exceed a voltage threshold, asynchronous signals are less efficient than synchronous signals. However, taking axonal propagation times into account, synchronous signals may not coincide significantly at the post-synaptic neuron contrary to asynchronous or polychronous signals. The term polychronous was first introduced in 2006, by E. Izhichevitch in {[}\protect\hyperlink{ref-SM9G0xBK}{62}{]}. He defines this term after pointing out a particular organization of the neurons of his spiking artificial neural network. The network is characterized by a timing-dependant learning rule for weights (STDP) and by fixed conduction delays between neurons. Due to the interplay between the delays and STDP, the spiking neurons spontaneously self-organize into groups and generate patterns of stereotypical polychronous activity, i.e.~exhibit reproducible time-locked but not synchronous firing patterns. The neurons composing a group discharge at different times, but due to delays, the spikes reach the postsynaptic neuron at the same time. This synchrony leads to the summation of the excitatory post-synaptic potentials evoked by each spike and thus to the crossing of the voltage threshold and to the discharge of a spike. According to the STDP rule, the neurons involved in this activity will see their weight of synaptic connection increase and thus, constitute a polychronous group. Interestingly, thanks to the fact that a neuron can be involved in different polychronous groups, the number of coexisting polychronous groups far exceeds the number of neurons in the network, resulting in an unprecedented memory capacity of the system.

@@

\hypertarget{detecting-precise-spiking-motifs-in-biological-raster-plots}{%
\section{Detecting precise spiking motifs in biological raster plots}\label{detecting-precise-spiking-motifs-in-biological-raster-plots}}

\{sec:detection\}

\hypertarget{spike-distances}{%
\subsection{spike distances}\label{spike-distances}}

J. D. Victor and K. P. Purpura, ``Nature and precision of temporal coding in visual cortex: a metric-space analysis,'' J. Neurophysiol., vol.~76, pp.~1310--1326, Aug.~1996.

>>>>

\section{Introduction: Importance of precise spike timings in the brain}\label{introduction-importance-of-precise-spike-timings-in-the-brain}
\subsection{Ultra fast neural codes for ultra fast vision}\label{ultra-fast-neural-codes-for-ultra-fast-vision}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{https://cdn-learn.adafruit.com/assets/assets/000/086/572/original/circuitpython_Thorpe-FabreThorpe-Neuroscience-Seeking-categories-in-the-brain-monkey-visual-latency-estimate.jpg}
\caption{From {[}\protect\hyperlink{ref-12dWDpY61}{6}{]}}\label{fig:thorpe}
\end{figure}

\subsection{How timing encodes analogous profile}\label{how-timing-encodes-analogous-profile}
An important characteristic of neuronal information is that it consists mainly in the transfer of action potentials, or spikes, which consist of brief impulses that propagate along the axons of neurons. These have the particularity of being essentially binary in their amplitude (that is to say, they are prototypical, all or nothing). An important consequence of the speed of processing is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only 120 ms consists of about ten processing stages following the ``forward'' pathways of the visual system, then this imposes that the processing in a single area is performed with a reduced number of spikes. Neurophysiologists tipcally use firing sequences to characterize the activity of neurons using different statistics on their individual timing {[}\protect\hyperlink{ref-zZUVkTfZ}{7}{]} but also the dependance across neurons {[}\protect\hyperlink{ref-13T2Gne9z}{8}{]}.

@@

\begin{figure}
\centering
\includegraphics{images/roc.jpg}
\caption{Latency coding. An input analog profile is encoded in latencies: the higher the contrast, the shorter the latency. In this example, one generates at most one spike per neuron. From {[}\protect\hyperlink{ref-YX5wqWH4}{16}{]}}\label{fig:roc}
\end{figure}

@@

\subsection{From synfire chains\ldots{}}\label{from-synfire-chains}
The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate is highly irregular {[}\protect\hyperlink{ref-sYHyiGsP}{26}{]}, which makes it inconsistent with temporal integration of random EPSPs. It has also been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible, where the first spike is aligned but the subsequent spike times tend to diffuse for independent repetitions of the stimulation {[}\protect\hyperlink{ref-XsHCA4IZ}{15}{]} (see Figure {[}\ref{fig:mainen}{]}). When driven by a highly dynamic \emph{frozen} noise, the output spikes are highly reproducible. This is consistent with the differential role of different stimulus frequencies on the reliability of spike timing reported in {[}\protect\hyperlink{ref-1AbvCTG03}{27}{]} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''

\begin{figure}
\centering
\includegraphics{https://camo.githubusercontent.com/77f4a48a0fe8f98ffab3c1f21dca1e74286b14d47c17120ee9627f3351fcd5e7/687474703a2f2f692e737461636b2e696d6775722e636f6d2f69786e727a2e706e67}
\caption{The timing of the spikes produced following the repetition of a step stimuli is less reproducible than that to a noisy stimulus. While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} nois at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this \href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{https://brian2.readthedocs.io/en/stable/_images/frompapers.Diesmann_et_al_1999.1.png}
\caption{Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{using Brian}. The model consists of 10 groups of 100 neurons each. A pulse with a given jitter is generated in the first group which here generates a pulse after a certain processing delay in the second group with a lower jitter. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{https://www.frontiersin.org/files/Articles/39056/fnins-07-00011-r2/image_m/fnins-07-00011-g004.jpg}
\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background is only utilized in the original model, where it is implemented as random Gaussian current. For the presented hardware implementation it has been omitted due to network size constraints. As compensation for missing background stimuli, the resting potential was increased to ensure a comparable excitability of the neurons. The state space gives the duration of synfire chains as a function of the protocol's parameters and compares (C) CPU to (D) neuromorphic implementations.}\label{fig:pynn}
\end{figure}

@@

\subsubsection{.. to traveling waves \ldots{}}\label{to-traveling-waves}
TODO : {[}\protect\hyperlink{ref-zNDCo9Hy}{49}{]}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{https://media.springernature.com/lw685/springer-static/image/art\%3A10.1038\%2Fs41467-021-26175-1/MediaObjects/41467_2021_26175_Fig1_HTML.png}
\caption{a Spike raster for repeated presentations (N = 40) of a high-contrast (10\% Michelson contrast) drifting Gabor recorded from area MT of a fixating marmoset (stimulus-onset, red line; mean response, blue line). A single-trial LFP trace is plotted in gray, and the average LFP response is plotted in black. The relative power between baseline (−200 ms to stimulus-onset) and evoked fluctuations (stimulus-onset + 50--250 ms) significantly favored the evoked response (right panel; N = 110 trials; median = 1.89 dB, p = 0.000019, two-tailed Wilcoxon's ranked-sum test). b Same as in (a), but for a low contrast stimulus (\textless2\% Michelson contrast). The relative power between baseline and evoked LFP fluctuations was not statistically different from parity (median = 1.23 dB, p = 0.087 two-tailed Wilcoxon's ranked-sum test). c An example of spontaneous LFP fluctuations structured as a traveling wave recorded from a spatially distributed multielectrode array in marmoset area MT. d Histogram of spontaneous spike probability as a function of the generalized phase of the LFP during fixation. e The average evoked response to low contrast stimuli was stronger when a more excitable phase (±π rad) of a spontaneous traveling wave aligned with the retinotopic location of the target (aligned, green line) as compared to when a less excitable phase (0 rad) was aligned (unaligned, purple line; N = 43 wave and non-wave trials; shaded region SEM; p = 0.0000015 two-tailed Wilcoxon's rank-sum test). Data for panels c--e modified from Davis et al.17 with permission.}\label{fig:davis}
\end{figure}

@@

\subsubsection{\texorpdfstring{Spike Synchronization and Rate Modulation Differentially Involved in Motor Cortical Function {[}\protect\hyperlink{ref-wVZpQvSk}{42}{]}}{Spike Synchronization and Rate Modulation Differentially Involved in Motor Cortical Function {[}42{]}}}\label{spike-synchronization-and-rate-modulation-differentially-involved-in-motor-cortical-function-doi10.1126science.278.5345.1950}
It is now commonly accepted that planning and execution of movements are based on distributed processing by neuronal populations in motor cortical areas {[}\protect\hyperlink{ref-yLVM9IXi}{57}{]}. It is less clear, though, how these populations organize dynamically to cope with the momentary computational demands. Simultaneously recorded activities of neurons in the primary motor cortex of monkeys during performance of a delayed-pointing task exhibited context-dependent, rapid changes in the patterns of coincident action potentials. Accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.

@@

\subsubsection{TODO \ldots{} and to synfire braids and polychronous groups}\label{todo-and-to-synfire-braids-and-polychronous-groups}
\begin{itemize}
\tightlist
\item
  {[}\protect\hyperlink{ref-HjDb1gnq}{59}{]} : from synfire chains to Synfire braids
\end{itemize}

@@

\subsection{Are there precise heterosynaptic spiking motifs in the brain?}\label{are-there-precise-heterosynaptic-spiking-motifs-in-the-brain}
Currently, there is a consensus for rate-based coding models in computational and biological neuroscience. Nevertheless, there is a substantial literature in neurobiology indicating that brain dynamics often organize into stereotyped sequences (like synfire chains {[}\protect\hyperlink{ref-Mf6h1KNY}{63}{]}, packets {[}\protect\hyperlink{ref-2GovoaYw}{64}{]} or hippocampal sequences {[}\protect\hyperlink{ref-SVfasyCp}{65},\protect\hyperlink{ref-PUNGVou}{66}{]} and on the role of such precise spike timing in downstream information transfer and coding {[}\protect\hyperlink{ref-19fAM6hcv}{67},\protect\hyperlink{ref-13dQ0R4UM}{68},\protect\hyperlink{ref-MfmcMsqA}{69}{]}. This is for instance relevant in sensory pathways in vision {[}\protect\hyperlink{ref-1EHZgcKyL}{70}{]}, audition {[}\protect\hyperlink{ref-FlKJw2VL}{71}{]}, olfaction {[}\protect\hyperlink{ref-AhSRXBJ2}{72}{]}{]} or touch {[}\protect\hyperlink{ref-gDinei4a}{73}{]}. In particular, one theoretical viewpoint considers synfire braids (Bienenstock, 1995), where a precise sequential motif of spikes will synchronize as it reaches the soma of a neuron for which synaptic delays are adequately tuned. In particular, computational modeling shows that at the scale of neurons, an efficient neural code can emerge where spike times are organized in prototypical, precise temporal motifs (Izhikevich, 2006) which he defined as polychronous groups.

@@

\begin{figure}
\centering
\includegraphics{https://www.pnas.org/cms/10.1073/pnas.1718518116/asset/ab9a48f6-b28f-40c1-978e-6f6dbec39b98/assets/graphic/pnas.1718518116fig01.jpeg}
\caption{Mixed distance and duration representation in CA1. (A) Calcium fluorescence (heatmap) of CA1 neurons participating to run sequences in consecutive imaging sessions. Cells have been selected and ordered with respect to their activity in the first imaging session (Top). The black line represents speed of the mouse.}\label{fig:haimerl}
\end{figure}

@@

\subsection{Cortical songs}\label{cortical-songs}
\begin{itemize}
\tightlist
\item
  Ikegaya and colleagues {[}\protect\hyperlink{ref-Mf6h1KNY}{63}{]} worked on spontaneous activity in vitro and in vivo. They demonstrated that in cortical activity, we can find a repetition of several motifs. In PSCs, but also in spike activity. These sequences repeat after minutes and have a precise spatio temporal structure with a ms precision. They can be specific of a particular layer or colomn, are synchronized with network activity oscillation and can involve several cells. They also demonstated that these sequences can form supersequences : the cortical songs. It consist of the assembly of several sequences which repeat in a specific order with a compressed timing.
\item
  ``We find precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons in vivo and in vitro. These patterns repeat after minutes, maintaining millisecond accuracy.''
\item
  Precisely repeating motifs of spontaneous synaptic activity in slices: duration around 1s +/- .5 s. Some events in motifs are of similar size but sometimes absent - better described by Bernouilli than SE (and covariance)
\item
  \emph{in vivo} spontaneous activity also reveals repeating sequences. About 3000 sequences, each involving 3-10 cells out of about 900, and last up to 3 seconds
\item
  topography: ``Sequences had specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster (Fig. 4, A and B, and fig.~S3B). (\ldots) Therefore, repeating temporal patterns of activation (\ldots) were associated with a structured spatial organization of the neurons that formed them.''
\item
  ``Cortical songs: modular assemblies of repeated sequences'': hierarchical detection.
\item
  in cotical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)
\end{itemize}

\begin{figure}
\centering
\includegraphics{images/Ikegaya2004zse0150424620001.jpeg}
\caption{Fig. 1. from {[}\protect\hyperlink{ref-Mf6h1KNY}{63}{]} Repeated motifs of spontaneous synaptic activity in vitro and in vivo. (A) Repeated motifs of intracellular activity from layer 5 pyramidal neurons in slices. Panels show segments (red) of the same voltage-clamp recording from the same cell repeating seconds or minutes after the initial occurrence (blue). Arrows indicate timings of repeated PSCs. (i) Upper trace: low--temporal resolution display of spontaneous activity of a neuron. Lower traces: higher resolution display of the repeated motif at indicated regions of the trace (a to c). (ii) Example of a longer motif. (B) Three repetitions of a motif. The top traces show the motifs superimposed on each other (blue, green, and red), the middle traces show these same traces individually, and the bottom traces show temporally magnified regions of the motifs (a to c). (C) Repeated sequences of intracellular current-clamp recordings in vivo. Two (i) and three (ii) repetitions of motifs are shown. Shuffle tests were performed on traces (i), a to c, yielding significantly fewer repeats (fig.~S2, P \textless{} 0.02). In (i), the blue trace is shifted --2.75 mV; in (ii), the blue trace is shifted --1.58 mV, and the green +0.79 mV.}\label{fig:Ikegaya2004}
\end{figure}

@@

\subsection{outline}\label{outline}
A natural candidate to use these precise temporal patterns in the brain is to use Spiking Neural Networks (SNNs) {[}\protect\hyperlink{ref-DUpFdm8C}{78}{]}. The approach which is at present most prominent in the SNNs community is to use existing algorithms from machine learning and to adapt them to the specificity of spiking architectures. One such example is to adapt the successes of deep learning algorithms and to transfer the back-propagation algorithm to SNNs, for instance with a surrogate gradient. This approach is quite successful, and SNNs approach in some case the performance of Deep Learning algorithms, for instance on the N-MNIST dataset for categorizing digits in a stream of events. However, most biological neural systems use spikes and are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption. There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs. Our approach will be to focus on the temporal representation of information directly. In particular, our objective is to fully exploit the capacity of spiking neurons to detect synchronous patterns.

@@

@@

\begin{figure}
\centering
\includegraphics{https://laurentperrinet.github.io/grant/anr-anr/event_driven_computations.png}
\caption{A miniature, event-based ATIS sensor. Contrary to a classical frame-based camera for which a full dense image representation is given at discrete, regularly spaced timings, the event-based camera provides with events at the micro-second resolution. These are sparse as they represent luminance increments or decrements (ON and OFF events, respectively).}\label{fig:silicon_retina}
\end{figure}

@@

\section{Theoretical models for the detection of precise heterosynaptic spiking motifs}\label{theoretical-models-for-the-detection-of-precise-heterosynaptic-spiking-motifs}
\#sec:detection

@@

\subsection{Izhikevitch's polychronization model}\label{izhikevitchs-polychronization-model}
Considering the number of spikes required to exceed a voltage threshold, asynchronous signals are less efficient than synchronous signals. However, taking axonal propagation times into account, synchronous signals may not coincide significantly at the post-synaptic neuron contrary to asynchronous or polychronous signals. The term polychronous was first introduced in 2006, by E. Izhichevitch in {[}\protect\hyperlink{ref-SM9G0xBK}{62}{]}. He defines this term after pointing out a particular organization of the neurons of his spiking artificial neural network. The network is characterized by a timing-dependant learning rule for weights (STDP) and by fixed conduction delays between neurons. Due to the interplay between the delays and STDP, the spiking neurons spontaneously self-organize into groups and generate patterns of stereotypical polychronous activity, i.e.~exhibit reproducible time-locked but not synchronous firing patterns. The neurons composing a group discharge at different times, but due to delays, the spikes reach the postsynaptic neuron at the same time. This synchrony leads to the summation of the excitatory post-synaptic potentials evoked by each spike and thus to the crossing of the voltage threshold and to the discharge of a spike. According to the STDP rule, the neurons involved in this activity will see their weight of synaptic connection increase and thus, constitute a polychronous group. Interestingly, thanks to the fact that a neuron can be involved in different polychronous groups, the number of coexisting polychronous groups far exceeds the number of neurons in the network, resulting in an unprecedented memory capacity of the system.

@@

\section{Detecting precise spiking motifs in biological raster plots}\label{detecting-precise-spiking-motifs-in-biological-raster-plots}
\{sec:detection\}

\subsection{spike distances}\label{spike-distances}
J. D. Victor and K. P. Purpura, ``Nature and precision of temporal coding in visual cortex: a metric-space analysis,'' J. Neurophysiol., vol.~76, pp.~1310--1326, Aug.~1996.
